A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-11-10 jour/jour.tex
Some clarifications

Let $\tEc$ denote the set of $\Sig$-terms occurring in $\Ec$.
At the end of the first phase, $\Ec$ induces an equivalence
relation over $\tEc$ such that two terms $t$ and $u$ are equivalent if and
only if $t \teq u \in \Ec$.
Thus, we can regard $\Ec$ as a set of
equivalence classes of terms. For a term $t \in \tEc$, we write $\ec{t}$ to
denote the equivalence class of $t$ in $\Ec$.

@@

The mapping $\Val$ is defined as follows.
%
With each equivalence class $\ec{u}$, we associate a fresh variable $\Varec{u}$
of the same type as $u$
%not occurring in $\tEc$,
and set $\Val\ec{u} := {\Varec{u}}$. Because $\Varec{u}$ is unspecified,
this indicates that there are initially no
constraints on the values for any equivalence class $\ec{u}$. The mapping $\Val$
is refined by applying the following unfolding rule exhaustively:\strut
\[
\hbox{\(
\inferrule{
  {\Varec{u}} \in \FV( \Val )
  \betweenantes
  \const{C}( t_1, \ldots, t_n ) \in \ec{u}
  \betweenantes
  \const{C} \in \Ctr
}{
  \Val := \Val [{\Varec{u}} \mapsto \MU {\Varec{u}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} )]
}
\)
%\rn{Unfold}
}
\]
$\FV( \Val )$ denotes the set of free variables occurring in $\Val$'s range,
and $\Val[x \mapsto t]$ denotes the \emph{variable-capturing} substitution of $t$ for
$x$ in $\Val$'s range. It is easy to see that the height of terms produced as a
result of the unfolding is bounded by the number of equivalence classes of
$\Ec$, and thus the construction of $\Val$ will terminate.

@@

\begin{examplex}
Let $\Ec = \{ x \teq \const{S}(x),\; y \teq \const{S}(\const{S}(y)) \}$.
After phase~1, the equivalence classes %of $\Ec$
are
$\{ x,\, \const{S}(x) \}$,
$\{ y,\, \const{S}(\const{S}(y)) \}$,
and $\{ \const{S}(y) \}$.
Constructing $\Val$ yields
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x}) &
\kern0.25em %% TYPESETTING
\Val \ec{y} & \,=\,
\MU {\Varec{y}}.\; \const{S}(\smash{\MU {\VAREC{\const{S}(y)}}}.\; \const{S}( {\Varec{y}} ) )
\kern0.25em %% TYPESETTING
 &
\Val \ec{\const{S}(y)} & =
\MU {\VAREC{\const{S}(y)}}.\; \const{S}(\MU {\Varec{y}}.\; \const{S}( {\VAREC{\const{S}(y)}} ) )
\end{align*}
Since $\Val \ec{y} \aequiv \Val \ec{\const{S}(y)}$,
%for the renaming
%$\{ {\Varec{y}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{y}} \}$,
the \rn{Unique} rule applies to derive $y \teq \const{S}(y)$.
%Intuitively, this equality holds, since the values of $y$ and $\const{S}(y)$
%are equivalent.
%
At this point, phase~1 is activated again, % to XXX the equivalence classes,
yielding
$\{ x,\, \const{S}(x) \}$ and
$\{ y,\, \const{S}(y),\allowbreak\, \const{S}(\const{S}(y)) \}$.
The mapping $\Val$ is updated accordingly:
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x})
&
\Val \ec{y} & \,=\, \MU {\Varec{y}}.\; \const{S}({\Varec{y}})
\end{align*}
Since $\Val \ec{x} \aequiv \Val \ec{y}$,
%for the renaming
%$\{ {\Varec{y}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{y}} \}$,
\rn{Unique} can finally derive $x \teq y$.
\xend
\end{examplex}

@@

The implementation is able to produce models for satisfiable inputs.

[TODO]

>>>>

Let $\tEc$ denote the set of $\Sig$-terms occurring in $\Ec$.
At the end of the first phase, $\Ec$ induces an equivalence
relation over $\tEc$ such that two terms $t$ and $u$ are equivalent if and
only if $t \teq u \in \Ec$.
Thus, we can regard $\Ec$ as a set of
equivalence classes of terms. For a term $t \in \tEc$, we write $\ec{t}$ to
denote the equivalence class of $t$ in $\Ec$.
Moreover, at the end of this phase, each equivalence class $\ec{t}$ contains (at most) 
one constructor term that is unique up to congruence.
Thus, in the subsequent phases, 
when considering the case that $\ec{t}$ contains a constructor term,
it is enough for our purposes to select an arbitrary constructor term $\const{C}( \vec t\, ) \in \ec{t}$ among these.

@@

The mapping $\Val$ is defined as follows.
%
With each equivalence class $\ec{u}$, we associate a fresh variable $\Varec{u}$
of the same type as $u$. %not occurring in $\tEc$,
%AJR : this sentence is necessary, since we are abusing the \Varec notation a bit
For a term $t \in \tEc$, 
we write $\Varec{t}$ to denote the variable associated with equivalence class $\ec{t}$.
Initially, we set $\Val\ec{u} := {\Varec{u}}$ for each equivalence class $\ec{u}$. Because $\Varec{u}$ is unspecified,
this indicates that there are initially no
constraints on the values for any equivalence class $\ec{u}$. 
The mapping $\Val$
is refined by applying the following unfolding rule exhaustively:\strut
\[
\hbox{\(
\inferrule{
  {\Varec{u}} \in \FV( \Val )
  \betweenantes
  \const{C}( t_1, \ldots, t_n ) \in \ec{u}
  \betweenantes
  \const{C} \in \Ctr
}{
  \Val := \Val [{\Varec{u}} \mapsto \MU {\Varec{u}}.\; \const{C}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} )]
}
\)
%\rn{Unfold}
}
\]
$\FV( \Val )$ denotes the set of free variables occurring in $\Val$'s range,
and $\Val[x \mapsto t]$ denotes the \emph{variable-capturing} substitution of $t$ for
$x$ in $\Val$'s range. It is easy to see that the height of terms produced as a
result of the unfolding is bounded by the number of equivalence classes of
$\Ec$, and thus the construction of $\Val$ will terminate.

@@

\begin{examplex}
Let $\Ec = \{ x \teq \const{S}(x),\; y \teq \const{S}(\const{S}(y)) \}$.
After phase~1, the equivalence classes %of $\Ec$
are
$\{ x,\, \const{S}(x) \}$,
$\{ y,\, \const{S}(\const{S}(y)) \}$,
and $\{ \const{S}(y) \}$.
Constructing $\Val$ yields
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x}) &
\kern0.25em %% TYPESETTING
\Val \ec{y} & \,=\,
\MU {\Varec{y}}.\; \const{S}(\smash{\MU {\VAREC{\const{S}(y)}}}.\; \const{S}( {\Varec{y}} ) )
\kern0.25em %% TYPESETTING
 &
\Val \ec{\const{S}(y)} & =
\MU {\VAREC{\const{S}(y)}}.\; \const{S}(\MU {\Varec{y}}.\; \const{S}( {\VAREC{\const{S}(y)}} ) )
\end{align*}
Since $\Val \ec{y} \aequiv \Val \ec{\const{S}(y)}$,
%for the renaming
%$\{ {\Varec{y}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{y}} \}$,
the \rn{Unique} rule applies to derive $y \teq \const{S}(y)$.
%Intuitively, this equality holds, since the values of $y$ and $\const{S}(y)$
%are equivalent.
%
At this point, phase~1 is activated again, % to XXX the equivalence classes,
yielding
$\{ x,\, \const{S}(x) \}$ andwill
$\{ y,\, \const{S}(y),\allowbreak\, \const{S}(\const{S}(y)) \}$.
The mapping $\Val$ is updated accordingly:
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x})
&
\Val \ec{y} & \,=\, \MU {\Varec{y}}.\; \const{S}({\Varec{y}})
\end{align*}
Since $\Val \ec{x} \aequiv \Val \ec{y}$,
%for the renaming
%$\{ {\Varec{y}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{y}} \}$,
\rn{Unique} can finally derive $x \teq y$.
\xend
\end{examplex}

@@

The implementation is able to produce models for satisfiable inputs.
As described in Section~\ref{sec:termination-and-correctness},
given a saturated set $\Ec$, we may construct a map $\ValC$ from variables (corresponding to the equivalence classes of $\Ec$) to closed normal $\MU$-terms.
%In practice, this requires a fair enumerator for each (co)datatype $\tau$.

%Recall that to assign a value $\MU {\Varec{x}}.\; t$ to variable

%We do not explicitly construct the set $\BAD_{\smash{\J}}^x( \Val )$.
%Instead,
